Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-08-14 analysis/paper/paper.Rmd
need to add: Figure 2: Map showing the largest section of excavation areas at Kiwulan

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
```

@@

\Although mountains isolated Yilan by land, the sea provided a way through which indigenous people in Yilan could interact with groups elsewhere, such as indigenous groups, distant traders such as the Han people of China, and Europeans. Large-scale and frequent trading activities brought a wide variety of objects into indigenous communities in Yilan. \The indigenous people of Yilan established their own exchange network that has been described as “inter-insular trade” by @Chen2005 [p. 12] to refer to small-scale regional trade between China and Taiwan on an irregular basis. One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads. These are one of the most abundant types of foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as social signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

@@

```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text", out.width= "400px"}
# need to add the locations of Keelung, Tamsui, and hoping dao
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.png"))
```

@@

```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number. Each square is 4 x 4 m", out.width= "70%", echo=FALSE}
# add to the map and scale and north arrow. Check it out after knitting 
knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.png"))
```

Kiwulan (Figure \@ref(fig:KWL-map)) is located at northern Yilan and was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 excavation squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan", out.width='60%'}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories

@@

ggplot(ornaments_count,
       aes(Categories, n)) +
  geom_col() +
  coord_flip() +
  labs(y = "Frequency",
       x = "Categories") +
  theme_minimal(base_size = 14)
```

@@

# histogram facetting for three periods
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(x = "categories", y = "Frequency") +
  theme_minimal(base_size = 12)
```

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequency for the major class of ornaments between different time periods at Kiwulan. The result of chi-square test for the number of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = `r chisq_test_result_p_value`) shows that there is a non-random difference between the three time periods. It indicates that most ornament types were present before European contact. Ornament frequencies reached a peak during European contact and then dropped during the Chinese contact period, especially for golden beads. This trend can be also seen on other ornaments including agate beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European contact period and a further decrease in the Chinese contact period.

@@

agate_bead_shape_plot <- 
ggplot(agate_bead_shape) +
  geom_col(aes(reorder(Type, n), n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "",
       x = "Agate bead") +
  theme_minimal(base_size = 14) +
  theme(strip.background = element_blank(), 
        strip.text = element_blank())
```

@@

ring_shape_plot <-
ggplot(ring_shape) +
  geom_col(aes(Type, n)) +
  coord_flip() +
  labs(y = "",
       x = "Metal ring") +
  facet_wrap(~ period) +
  theme_minimal(base_size = 14) 
```

@@

bell_shape_plot <-
ggplot(bell_shape) +
  geom_col(aes(reorder(Type,n), n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "",
       x = "Bell") +
  theme_minimal(base_size = 14) +
  theme(strip.background = element_blank(), 
        strip.text = element_blank())
```

@@

glass_bead_plot <-
ggplot(glass_bead_shape) +
  geom_col(aes(Type, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "Frequency",
       x = "Glass bead") +
  theme_minimal(base_size = 14) +
  theme(strip.background = element_blank(), 
        strip.text = element_blank())
```

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across contact periods for metal rings, agate beads, bells, and glass beads."}
library(cowplot)
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
```

@@

The distribution of frequency for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p-value = `r spearman_result_p_value`). This indicates that the increases in diversity can be explained by the effects of culture contact instead of effects of sample size. According to the frequency, agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European contact period. The greater varieties for agate beads and metal rings might indicate multiple origins due to participation in global trade networks stimulated by the Europeans. In contrast, copper bells and glass beads have less variety, but glass beads have a larger number because they were usually found in a cluster that indicates the original use as strings or necklaces. The common shape for copper bells is the large shape with a wide variety of human faces as a motif, while the common shape for glass beads is a bead with length less than 1 cm that are often identified as Indo-Pacific beads, the most common type that had been widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

# plot spatial pattern for all ornaments
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))
```

```{r, plot-spatial-pattern-each, fig.cap= "Spatial pattern for ornament class by time periods, only those types with more than 5 pieces are shown here"}

# filter golden bead
ornaments_period_join_gold <-
  ornaments_period_join %>% 
  filter(Categories %in% 'Golden bead')

# plot by period
dis_golden <-
ggplot(ornaments_period_join_gold) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_gold, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Golden Beads",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

# filter glass bead
ornaments_period_join_gold <-
  ornaments_period_join %>% 
  filter(Categories %in% 'Glass bead')

# plot by period
dis_glass <-
  ggplot(ornaments_period_join_gold) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_gold, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Glass Beads",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

# filter agate bead
ornaments_period_join_gold <-
  ornaments_period_join %>% 
  filter(Categories %in% 'Agate bead')

# plot by period
dis_agate <-
ggplot(ornaments_period_join_gold) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_gold, 
          aes(geometry = geometry, 
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Agate Beads",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

plot_grid(dis_golden,
          dis_glass,
          dis_agate, 
          align = "v",
          nrow = 3)
```

@@

# kernel density estimation with contour
kernel_tmp <- density(ornaments_point_ppp_tmp)
kernel_to_plot <- 
  plot(kernel_tmp, ...)
  contour(kernel_tmp, add = TRUE)

@@

```{r plot-all-densities, fig.cap= "Kernal density map for ornaments by periods"}
par(mfrow = c(1, 3))
# apply function to each data frame in the list
titles <- names(ornaments_period_join_cen)

kernels_for_plotting <- 
  map2(ornaments_period_join_cen, 
       titles,
      ~density_analysis(.x, main = .y))
```

@@

```{r hypo-test-distribution, fig.cap= "Histogram of simulated ANN values with the blue line indicating the observed value"}

# make the blue lines thicker to be more visible
par(mfrow = c(3, 1))
invisible(
map2(kernels_for_plotting, 
    titles,
    ~invisible(nndist_hypo_testing(.x, 
                         main = .y)))
)
```

>>>>

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
# prepare for JAS:R
# https://www.elsevier.com/journals/journal-of-archaeological-science-reports/2352-409x/guide-for-authors#2000
# not exceed 6000 words excluding figures, tables, highlights 
image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
```

@@

Although mountains isolated Yilan by land, the sea provided a way through which indigenous people in Yilan could interact with groups elsewhere, such as indigenous groups, distant traders such as the Han people of China, and Europeans. Large-scale and frequent trading activities brought a wide variety of objects into indigenous communities in Yilan. The indigenous people of Yilan established their own exchange network that has been described as part of an "inter-insular trade" system [@Chen2005, p. 12] to refer to small-scale regional trade between China and Taiwan on an irregular basis. One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads. These are one of the most abundant types of foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as social signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

@@

```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text", fig.width=image_width_2_col_inch}
# need to add the locations of Keelung, Tamsui, and hoping dao
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.png"))
```

@@

```{r, KWL-sam-area, fig.cap = "Map showing the largest section of excavation areas at Kiwulan, and the distribution of forty squares sampled in this paper presented in red with square ID number. Each square is 4 x 4 m", fig.width=image_width_2_col_inch}
# add to the map and scale and north arrow. Check it out after knitting 
knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.png"))
```

Kiwulan is located at northern Yilan (Figure \@ref(fig:KWL-map)) and was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge . The total area includes eight open area sections of 262 excavation squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan", fig.width=image_width_1_col_inch}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories

@@

plot_ornaments_count <- 
ggplot(ornaments_count,
       aes(Categories, n)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(limits = c(0, 160)) +
  labs(y = "Frequency",
       x = "Categories") +
  theme_minimal(base_size = 10)

ggsave(plot = plot_ornaments_count,
       here("analysis", "figures",
       "plot-ornaments-count.png"),
       width = 90, 
       height = 40,
       units = "mm")

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-ornaments-count.png"))
```

@@

# histogram facetting for three periods
plot_ornaments_count_three_periods <- 
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(x = "Categories",
       y = "Frequency") +
  theme_minimal(base_size = 12)

ggsave(plot = plot_ornaments_count_three_periods,
       here("analysis", "figures",
       "plot-ornaments-count-three-periods.png"),
       width = 190, 
       height = 80,
       units = "mm")

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-ornaments-count-three-periods.png"))
```

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequency for the major class of ornaments between different time periods at Kiwulan. The result of chi-square test for the number of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p = $`r round(chisq_test_result_p_value, 14)`$) shows that there is a non-random difference between the three time periods. It indicates that most ornament types were present before European contact. Ornament frequencies reached a peak during European contact and then dropped during the Chinese contact period, especially for golden beads. This trend can be also seen on other ornaments including agate beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European contact period and a further decrease in the Chinese contact period.

@@

base_size_multi_plot <- 10

agate_bead_shape_plot <- 
ggplot(agate_bead_shape) +
  geom_col(aes(reorder(Type, n), n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "",
       x = "Agate bead") +
  theme_minimal(base_size = base_size_multi_plot) +
  theme(strip.background = element_blank(), 
        strip.text = element_blank())
```

@@

ring_shape_plot <-
ggplot(ring_shape) +
  geom_col(aes(Type, n)) +
  coord_flip() +
  labs(y = "",
       x = "Metal ring") +
  facet_wrap(~ period) +
  theme_minimal(base_size = base_size_multi_plot) 
```

@@

bell_shape_plot <-
ggplot(bell_shape) +
  geom_col(aes(reorder(Type,n), n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "",
       x = "Bell") +
  theme_minimal(base_size = base_size_multi_plot) +
  theme(strip.background = element_blank(), 
        strip.text = element_blank())
```

@@

glass_bead_plot <-
ggplot(glass_bead_shape) +
  geom_col(aes(Type, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "Frequency",
       x = "Glass bead") +
  theme_minimal(base_size = base_size_multi_plot) +
  theme(strip.background = element_blank(), 
        strip.text = element_blank())
```

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across contact periods for metal rings, agate beads, bells, and glass beads."}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))

ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.png"),
       width = 190, 
       height = 150,
       units = "mm")

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.png"))
```

@@

The distribution of frequency for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture contact instead of effects of sample size. According to the frequency, agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European contact period. The greater varieties for agate beads and metal rings might indicate multiple origins due to participation in global trade networks stimulated by the Europeans. In contrast, copper bells and glass beads have less variety, but glass beads have a larger number because they were usually found in a cluster that indicates the original use as strings or necklaces. The common shape for copper bells is the large shape with a wide variety of human faces as a motif, while the common shape for glass beads is a bead with length less than 1 cm that are often identified as Indo-Pacific beads, the most common type that had been widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

# plot spatial pattern for all ornaments
plot_spatial_distribution_all_ornaments_by_period <- 
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

ggsave(plot = plot_spatial_distribution_all_ornaments_by_period,
       here("analysis", "figures",
       "plot-spatial-distribution-all-ornaments-by-period.png"),
       width = 190, 
       height = 50,
       units = "mm")

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-spatial-distribution-all-ornaments-by-period.png"))
```

```{r, plot-spatial-pattern-each, fig.cap="Spatial pattern for ornament class by time periods, only those types with more than 5 pieces are shown here"}

ornaments_period_join_selected_types <- 
  ornaments_period_join %>% 
  filter(Categories %in% c('Golden bead', 
                           'Glass bead',
                           'Agate bead'))

plot_spatial_distribution_some_ornaments_by_period <- 
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_selected_types, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_grid(rows = vars(Categories),
             cols = vars(period)) +
  labs(title = "",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 8))

ggsave(plot = plot_spatial_distribution_some_ornaments_by_period,
       here("analysis", "figures",
       "plot-spatial-distribution-some-ornaments-by-period.png"),
       width = 190, 
       height = 130,
       units = "mm")

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-spatial-distribution-some-ornaments-by-period.png"))

```

@@

# kernel density estimation with contour
kernel_tmp <- density(ornaments_point_ppp_tmp)
kernel_to_plot <- 
  plot(kernel_tmp, main = "")
  contour(kernel_tmp, add = TRUE)
  title(..., line = -2, adj = 0.4)

@@

```{r plot-all-densities, fig.cap= "Kernal density map for ornaments by periods"}

# apply function to each data frame in the list
titles <- names(ornaments_period_join_cen)

png(
  here("analysis", "figures",
       "plot-kde-maps.png"), 
    width = 190,
    height = 63,
    units = "mm",
  res = 300)

par(mfrow = c(1, 3),
    oma = c(1,1,1,1) + 0.1,
    mar = c(0,0,1,1) + 0.1)

kernels_for_plotting <- 
  map2(ornaments_period_join_cen, 
       titles,
      ~density_analysis(.x, main = .y))

invisible(dev.off())

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-kde-maps.png"))
```

@@

```{r hypo-test-distribution, fig.cap= "Histograms of simulated ANN values with the blue line indicating the observed value"}

# make the blue lines thicker to be more visible
png(
  here("analysis", "figures",
       "plot-kde-ann-histograms.png"), 
    width = 90,
    height = 150,
    units = "mm",
  res = 300)
par(mfrow = c(3, 1))
invisible(
map2(kernels_for_plotting, 
    titles,
    ~invisible(nndist_hypo_testing(.x, 
                         main = .y))))
invisible(dev.off())

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-kde-ann-histograms.png"))
```
